---
title: "Quiz 3"
author: "Nzambuli Daniel"
date: "2024-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# QUESTION 1 (10 MARKS):

## A cell type can exist in four states: A growth state, a resting state, a reproduction state, or it can be dead. It starts in the growth state. From the growth state it moves to the resting state with a rate of 0.3 and to the reproduction state with a rate 0.1, or it dies with a rate of 0.1. From the resting state it moves to the growth state with a rate of 0.2, or it dies with a rate of 0.05. From the reproduction state it moves to the resting state at a rate 0.5, or it dies at a rate of 0.3.

$$
generatot\ matrix\ Q = \left[\begin{matrix}
-0.5 & 0.3 & 0.1 & 0.1 \\
0.2 & -0.25 & 0 & 0.05 \\
0 & 0.5 & -0.8 & 0.3 \\
0 & 0 & 0 & 0
\end{matrix}\right]
$$

> -   the sum of all rows is equal to `0`

Generator matrix

$$
P\ matrix = \left[\begin{matrix}
0 & 0.6 & 0.2 & 0.2\\
0.8 & 0 & 0 & 0.2\\
0 & 0.625 & 0 & 0.375\\
0 & 0 & 0 & 0
\end{matrix}\right]
$$

> -   main diagonal = 0
>
> -   the row sum is 1

```{r}
library(DiagrammeR)
graph.code <- "digraph {
  graph [rankdir = 'LR'];
  node [fontname = 'helvetica', shape = circle];
  
  # Define the nodes
  tab1 [label = 'Growth', fillcolor = '#48b100', style = filled];
  tab2 [label = 'Resting', fillcolor = '#906f00', style = filled];
  tab3 [label = 'Reproduction', fillcolor = '#6f9000', style = filled];
  tab4 [label = 'death', fillcolor = '#ff0000', style = filled];
  
  # Edge definitions with the node IDs
  tab1 -> tab2 [label = '0.6'];
  tab1 -> tab3 [label = '0.2'];
  tab1 -> tab4 [label = '0.2'];
  tab2 -> tab1 [label = '0.8'];
  tab2 -> tab4 [label = '0.2'];
  tab3 -> tab2 [label = '0.625'];
  tab3 -> tab4 [label = '0.375'];
}"

grViz(graph.code)
```

> **Death** is an absorption state. Once a cell has died it can not come back to any of its previous states
>
> **Resting** a cell in resting state can only go into growth or death. However, the cell tends to go into a growth state more times than it dies.
>
> **Growth** a cell growing tends to go into a resting states $\frac {6}{10}$ times and has an equal likelihood of going to `reproduction` or `death` the rest of the time
>
> **Reproduction** a cell in the reproduction state can only go into resting and death states. The cell is likely to go to the resting state $\frac{5}{8}$ times.

### a) What is the expected lifetime of the cell? Compute the answer and describe in detail how such computations are done.

The lifespan of a cell is the time the cell takes to move from a given state to the absorbing state(**death**).

To find this the **fundamental matrix** needs to be found

> **Fundamental Matrix**
>
> -   A matrix of the states excluding the absorption state

$$
F\ (fundamental\ matrix) = \left[\begin{matrix}
-0.5 & 0.3 & 0.1\\
0.2 & -0.25 & 0\\
0 & 0.5 & -0.8
\end{matrix}\right]
$$

to find the absorption rate

$$
absorption\ time = -1 * \frac{1}{|F|} * \left[A^{adj}\right]^t
$$

this can also be done by

$$
absorption\ rate\ (N) = -1 * (F)^{-1}
$$

```{r}
f_mat = matrix(c(-0.5,0.3, 0.1, 0.2, -0.25, 0, 0, 0.5,-0.8), nrow = 3,byrow = TRUE)
colnames(f_mat) = c("growth", "resting", "reproduction")
rownames(f_mat) = c("growth", "resting", "reproduction")
f_mat
```

```{r}
# inverse of the matrix
inv_f = solve(f_mat)


# absorption of the matrix
abs_rate = -1 * inv_f

abs_rate
```

#### Time to death for cells

```{r}
rowSums(abs_rate)
```

> **From the question statement**
>
> all cells begin in the growth state therefore the lifespan of a cell is **12.261 units of time**

#### Interpretation

```{r}
grViz("digraph{
      graph [rankdir = 'LR'];
      node [format = 'helvetica', shape = 'circle'];
      
      # the nodes
      tab1 [label = 'Growth', fillcolor = '#35fe00', style = filled];
      tab2 [label = 'Death', fillcolor = '#fe3500', style = filled];
      
      
      # Arch
      tab1 -> tab2 [label = '12.26 units of time'];
  
      
}")
```

> It takes `12 units of time` from the growth stage to dying

```{r}
grViz("digraph{
      graph [rankdir = 'LR'];
      node [format = 'helvetica', shape = 'circle'];
      
      # the nodes
      tab1 [label = 'Resting', fillcolor = '#bcfe00', style = filled];
      tab2 [label = 'Death', fillcolor = '#fe3500', style = filled];
      
      
      # Arch
      tab1 -> tab2 [label = '13.81 units of time'];
  
      
}")
```

> It takes `12 units of time` from the resting stage to dying

```{r}
grViz("digraph{
      graph [rankdir = 'LR'];
      node [format = 'helvetica', shape = 'circle'];
      
      # the nodes
      tab1 [label = 'Reproduction', fillcolor = '#defe00', style = filled];
      tab2 [label = 'Death', fillcolor = '#fe3500', style = filled];
      
      
      # Arch
      tab1 -> tab2 [label = '9.88 units of time'];
  
      
}")
```

> It takes `10 units of time` from the reproduction stage to dying

------------------------------------------------------------------------

-   ::: {style="background-color: #4ac6f0;"}
    -   the cell has to move from **Resting** to **Growth** before it dies if it does not die immediately after transitioning from the resting state. This is reflected in the longer time it takes for the cell to die once it starts at the resting state

    -   A **Reproducing** cell dies the fastest. This can be as a result of multiple division and loss of material from within the cell leading to a faster deterioration

    -   **Resting** state can mean a preserved cell this may also account for the transition from preservation to initial growth and then reproduction with death as a finality. It could also account for the death as not all cells can recover from a preservation state.
    :::

------------------------------------------------------------------------

# Question 2

## Consider the Markov chain with state space E = {1, 2, 3} and transition matrix

$$
P = \left[\begin{matrix}
0.2 & 0.3 & 0.5\\
0.4 & 0.4 & 0.2\\
0&0&1
\end{matrix}\right]
$$

### a.) Using R program, draw a well labeled transitional plot of the Markov chain P

```{r}
grViz("digraph{
      graph [rankdir = 'LR'];
      node [format = 'helvetica', shape = 'circle'];
      
      # the nodes
      tab1 [label = 'A', fillcolor = '#35fe00', style = filled];
      tab2 [label = 'B', fillcolor = '#fe3500', style = filled];
      tab3 [label = 'c', fillcolor = '#0035fe', style = filled];

      
      
      # Arch
      tab1 -> tab1 [label = '0.2'];
      tab1 -> tab2 [label = '0.3'];
      tab1 -> tab3 [label = '0.5'];
      tab2 -> tab1 [label = '0.4'];
      tab2 -> tab2 [label = '0.4'];
      tab2 -> tab3 [label = '0.2'];
      tab3 -> tab3 [label = '1'];
  
      
}")   

```

> There is communication between state `A` and state `B`
>
> the state `c` is an **absorption state** once the states change to state c it keeps recurring within c but there is no chance of anything moving out of state c

### b.) Identify the recurrent states, transient states, and the absorbing states

The transient states are `a` and `b` as they communicate between each other in a cyclic manner

The absorbing state is state `c` as it only receives but does not output any value except to itself

The recurrent states are `a` , `b` and `c` as they have entry into the state and a self-communication

## Using R scripts determine the limiting transition probability of P

**Limiting transition probability** – is the steady state of a matrix `P`

```{r}
library(markovchain)

q2_p = matrix(c(0.2, 0.3, 0.5, 0.4, 0.4, 0.2, 0, 0, 1), nrow = 3, byrow = TRUE)
states = c('a', 'b', 'c')
colnames(q2_p) = states
rownames(q2_p) = states
q2_p
```

### Find the steady state

```{r}
markov_chain = new("markovchain", states = states,transitionMatrix = q2_p)
steady = steadyStates(markov_chain)
steady
```

The overall weight of any cost calculation based on the transition matrix P can be calculated as

$$
cost = 0a + 0b + c
$$

where `a` `b` and `c` are the individual costs in state a, b and c respectively
