---
title: "Quiz 3"
author: "Nzambuli Daniel"
date: "2024-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# QUESTION 1 (10 MARKS):

## A cell type can exist in four states: A growth state, a resting state, a reproduction state, or it can be dead. It starts in the growth state. From the growth state it moves to the resting state with a rate of 0.3 and to the reproduction state with a rate 0.1, or it dies with a rate of 0.1. From the resting state it moves to the growth state with a rate of 0.2, or it dies with a rate of 0.05. From the reproduction state it moves to the resting state at a rate 0.5, or it dies at a rate of 0.3. 

$$
Q\ matrix = \left[\begin{matrix}
-0.5 & 0.3 & 0.1 & 0.1 \\
0.2 & 0.25 & 0 & 0.05 \\
0 & 0.5 & -0.8 & 0.3 \\
0 & 0 & 0 & 0
\end{matrix}\right]
$$

> -   there is no transition rate for recurrence
>
> -   the sum of all rows is equal to `0`

Transition Matrix

$$
P\ matrix = \left[\begin{matrix}
0 & 0.6 & 0.2 & 0.2\\
0.8 & 0 & 0 & 0.2\\
0 & 0.625 & 0 & 0.375\\
0 & 0 & 0 & 0
\end{matrix}\right]
$$

> -   main diagonal = 0
>
> -   the row sum is 1

```{r}
library(DiagrammeR)
graph.code <- "digraph {
  graph [rankdir = 'LR'];
  node [fontname = 'helvetica', shape = circle];
  
  # Define the nodes
  tab1 [label = 'Growth', fillcolor = '#48b100', style = filled];
  tab2 [label = 'Resting', fillcolor = '#906f00', style = filled];
  tab3 [label = 'Reproduction', fillcolor = '#6f9000', style = filled];
  tab4 [label = 'death', fillcolor = '#ff0000', style = filled];
  
  # Edge definitions with the node IDs
  tab1 -> tab2 [label = '0.6'];
  tab1 -> tab3 [label = '0.2'];
  tab1 -> tab4 [label = '0.2'];
  tab2 -> tab1 [label = '0.8'];
  tab2 -> tab4 [label = '0.2'];
  tab3 -> tab2 [label = '0.625'];
  tab3 -> tab4 [label = '0.375'];
}"

grViz(graph.code)
```

> **Death** is an absorption state. Once a cell has died it can not come back to any of its previous states
>
> **Resting** a cell in resting state can only go into growth or death. However, the cell tends to go into a growth state more times than it dies.
>
> **Growth** a cell growing tends to go into a resting states $\frac {6}{10}$ times and has an equal likelihood of going to `reproduction` or `death` the rest of the time
>
> **Reproduction** a cell in the reproduction state can only go into resting and death states. The cell is likely to go to the resting state $\frac{5}{8}$ times.

### a) What is the expected lifetime of the cell? Compute the answer and describe in detail how such computations are done.

The lifespan of a cell is the time the cell takes to move from a given state to the absorbing state(**death**).

To find this the **fundamental matrix** needs to be found

> **Fundamental Matrix**
>
> -   A matrix of the states excluding the absorption state

$$
F\ (fundamental\ matrix) = \left[\begin{matrix}
-0.5 & 0.3 & 0.1\\
0.2 & -0.25 & 0\\
0 & 0.5 & -0.8
\end{matrix}\right]
$$

to find the absorption rate

$$
absorption\ time = -1 * \frac{1}{|F|} * \left[A^{adj}\right]^t
$$

this can also be done by

$$
absorption\ rate\ (N) = -1 * (F)^{-1}
$$

```{r}
f_mat = matrix(c(-0.5,0.3, 0.1, 0.2, -0.25, 0, 0, 0.5,-0.8), nrow = 3,byrow = TRUE)
colnames(f_mat) = c("growth", "resting", "reproduction")
rownames(f_mat) = c("growth", "resting", "reproduction")
f_mat
```

```{r}
# inverse of the matrix
inv_f = solve(f_mat)


# absorption of the matrix
abs_rate = -1 * inv_f

abs_rate
```

#### Time to death for cells

```{r}
rowSums(abs_rate)
```
